---
title: "Golden Health Index Regression: Full Reproduction"
subtitle: "Interactive Analysis Notebook for Reviewers"
author: "Generated from Paper Code"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = FALSE
)
```

# Overview

This notebook provides a complete walkthrough of the analyses presented in our paper on Golden Health Index (GHI) Regression. It is designed to be run by reviewers using **simulated data** that matches the structure and properties of the original NACC dataset.

## What This Notebook Does

1. **Setup**: Install dependencies and configure paths
2. **Data Exploration**: Create Table 1 (descriptive statistics)
3. **Model Fitting**: Fit the two-part model (GLM + LTRC)
4. **Model Analysis**: Generate Table 2 (coefficient estimates) and Figures 1-2
5. **Model Checking**: Comprehensive diagnostics for both model parts
6. **Sensitivity Analyses**: ADRC, temporal, and weighted analyses
7. **Simulation Study**: Small-scale demonstration

## Important Notes

- This notebook uses **simulated data** generated from fitted models
- Results will differ from the paper (which uses real NACC data), but the structure and methods are identical
- Computationally intensive sections (bootstrap variance) include warnings and alternatives
- All code can be run locally on a standard laptop
- **ADRC random effects analysis** requires TMB (Template Model Builder):
  - The `inst/paper_code/model/truncated_lmm.cpp` file will be automatically compiled
  - Requires a C++ compiler (Rtools on Windows, Xcode on Mac, g++ on Linux)
  - TMB installation: `install.packages("TMB")`

---

# 1. Setup {.tabset}

## Dependencies

First, let's install and load all required packages:

```{r install-packages, eval=FALSE}
# Install packages if needed
packages <- c(
  "readr", "dplyr", "ggplot2", "lubridate", "stringr", "patchwork",
  "fs", "data.table", "survival", "tictoc", "xtable", "purrr",
  "ResourceSelection", "car", "splines", "lme4", "sandwich", "lmtest",
  "truncreg", "statmod", "truncnorm", "TMB"
)

install.packages(setdiff(packages, rownames(installed.packages())))

# Install ltrc package (if not already installed)
if (!"ltrc" %in% rownames(installed.packages())) {
  devtools::install_github("srmatth/ltrc")
}
```

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(patchwork)
library(fs)
library(data.table)
library(survival)
library(tictoc)
library(xtable)
library(ResourceSelection)
library(car)
library(splines)
library(lme4)
library(sandwich)
library(lmtest)
library(purrr)

# Load ltrc if available
if (require("ltrc", quietly = TRUE)) {
  library(ltrc)
} else {
  warning("ltrc package not loaded. Some analyses will not run.")
}
```

## Configure Paths

Set up directory paths for data and results:

```{r configure-paths}
# Base reference age
A0 <- 65

# Data directory
DATA_DIR <- "~/Documents/repos/ghi/inst/extdata" # MAY NEED TO CHANGE THIS
if (!dir.exists(DATA_DIR)) {
  dir.create(DATA_DIR, recursive = TRUE)
}

# Results directory
RESULTS_DIR <- "~/Documents/repos/ghi/inst/paper_results"
if (!dir.exists(RESULTS_DIR)) {
  dir.create(RESULTS_DIR, recursive = TRUE)
}

# Bootstrap results directory (for variance estimation)
BOOTSTRAP_DIR <- "~/Documents/repos/ghi/inst/sim_res/NACC_mod_bs_with_race"
if (!dir.exists(BOOTSTRAP_DIR)) {
  dir.create(BOOTSTRAP_DIR, recursive = TRUE)
}

cat("✓ Directories configured\n")
cat("  Data:", DATA_DIR, "\n")
cat("  Results:", RESULTS_DIR, "\n")
cat("  Bootstrap:", BOOTSTRAP_DIR, "\n")
```

## Load Data

Load the simulated NACC dataset:

```{r load-data}
# Automatically use simulated data if real data not available
if (fs::file_exists(fs::path(DATA_DIR, "NACC_mod_dat.csv"))) {
  mod_dat <- read_csv(fs::path(DATA_DIR, "NACC_mod_dat.csv"))
  cat("✓ Loaded real NACC data\n")
} else if (fs::file_exists(fs::path(DATA_DIR, "NACC_mod_dat_simulated.csv"))) {
  mod_dat <- read_csv(fs::path(DATA_DIR, "NACC_mod_dat_simulated.csv"))
  cat("✓ Loaded simulated NACC data\n")
  cat("⚠ Note: Results will differ from paper (using simulated data)\n")
} else {
  stop("No data file found. Please ensure data directory is correct.")
}

# Display data structure
cat("\nDataset dimensions:", nrow(mod_dat), "observations,", ncol(mod_dat), "variables\n")
glimpse(mod_dat)
```

---

# 2. Data Exploration

This section reproduces **Table 1** from the main paper, showing descriptive statistics overall and stratified by APOE ε4 allele status.

## Overall Sample Characteristics

```{r table1-overall}
# Sample size
n_total <- nrow(mod_dat)
n_dementia <- sum(mod_dat$got_dementia)
n_no_dementia <- sum(mod_dat$got_dementia == 0)

# Age at study entry
age_entry_mean <- mean(mod_dat$age_at_study_entry)
age_entry_sd <- sd(mod_dat$age_at_study_entry)

# Age at death
age_death_mean <- mean(mod_dat$age_at_death)
age_death_sd <- sd(mod_dat$age_at_death)

# Age at dementia (among those with dementia)
age_dementia_mean <- mean(mod_dat$age_at_dementia[mod_dat$got_dementia == 1], na.rm = TRUE)
age_dementia_sd <- sd(mod_dat$age_at_dementia[mod_dat$got_dementia == 1], na.rm = TRUE)

# Sex
n_female <- sum(mod_dat$is_female)
pct_female <- mean(mod_dat$is_female) * 100

# Education
educ_mean <- mean(mod_dat$years_education)
educ_sd <- sd(mod_dat$years_education)

# Marital status
n_married <- sum(mod_dat$is_married)
pct_married <- mean(mod_dat$is_married) * 100

# Comorbidity
n_comorbidity <- sum(mod_dat$comorbidity)
pct_comorbidity <- mean(mod_dat$comorbidity) * 100

# Race
n_white <- sum(mod_dat$is_race_white)
pct_white <- mean(mod_dat$is_race_white) * 100
n_black <- sum(mod_dat$is_race_black)
pct_black <- mean(mod_dat$is_race_black) * 100
n_other <- sum(mod_dat$is_race_other)
pct_other <- mean(mod_dat$is_race_other) * 100

# APOE ε4 alleles
n_e4_0 <- sum(mod_dat$num_e4 == 0)
pct_e4_0 <- mean(mod_dat$num_e4 == 0) * 100
n_e4_1 <- sum(mod_dat$num_e4 == 1)
pct_e4_1 <- mean(mod_dat$num_e4 == 1) * 100
n_e4_2 <- sum(mod_dat$num_e4 == 2)
pct_e4_2 <- mean(mod_dat$num_e4 == 2) * 100

# Create summary table
table1_overall <- data.frame(
  Characteristic = c(
    "Sample size",
    "Age at study entry, mean (SD)",
    "Dementia status, n (%)",
    "  Developed dementia",
    "  Did not develop dementia",
    "Age at death, mean (SD)",
    "Age at dementia onset (if dementia), mean (SD)",
    "Female, n (%)",
    "Years of education, mean (SD)",
    "Married, n (%)",
    "≥1 Comorbidity, n (%)",
    "Race, n (%)",
    "  White",
    "  Black",
    "  Other",
    "APOE ε4 alleles, n (%)",
    "  0",
    "  1",
    "  2"
  ),
  Value = c(
    as.character(n_total),
    sprintf("%.1f (%.1f)", age_entry_mean, age_entry_sd),
    "",
    sprintf("%d (%.1f%%)", n_dementia, n_dementia/n_total*100),
    sprintf("%d (%.1f%%)", n_no_dementia, n_no_dementia/n_total*100),
    sprintf("%.1f (%.1f)", age_death_mean, age_death_sd),
    sprintf("%.1f (%.1f)", age_dementia_mean, age_dementia_sd),
    sprintf("%d (%.1f%%)", n_female, pct_female),
    sprintf("%.1f (%.1f)", educ_mean, educ_sd),
    sprintf("%d (%.1f%%)", n_married, pct_married),
    sprintf("%d (%.1f%%)", n_comorbidity, pct_comorbidity),
    "",
    sprintf("%d (%.1f%%)", n_white, pct_white),
    sprintf("%d (%.1f%%)", n_black, pct_black),
    sprintf("%d (%.1f%%)", n_other, pct_other),
    "",
    sprintf("%d (%.1f%%)", n_e4_0, pct_e4_0),
    sprintf("%d (%.1f%%)", n_e4_1, pct_e4_1),
    sprintf("%d (%.1f%%)", n_e4_2, pct_e4_2)
  )
)

knitr::kable(table1_overall, caption = "Table 1: Overall Sample Characteristics", align = "lr")
```

## Characteristics by APOE ε4 Status

```{r table1-stratified}
# Create stratified summaries
table1_stratified <- mod_dat %>%
  group_by(num_e4) %>%
  summarize(
    n = n(),
    age_entry = sprintf("%.1f (%.1f)", mean(age_at_study_entry), sd(age_at_study_entry)),
    n_dementia = sum(got_dementia),
    pct_dementia = sprintf("%.1f%%", mean(got_dementia) * 100),
    age_death = sprintf("%.1f (%.1f)", mean(age_at_death), sd(age_at_death)),
    age_dementia = sprintf("%.1f (%.1f)",
                          mean(age_at_dementia[got_dementia == 1], na.rm = TRUE),
                          sd(age_at_dementia[got_dementia == 1], na.rm = TRUE)),
    n_female = sum(is_female),
    pct_female = sprintf("%.1f%%", mean(is_female) * 100),
    educ = sprintf("%.1f (%.1f)", mean(years_education), sd(years_education)),
    n_married = sum(is_married),
    pct_married = sprintf("%.1f%%", mean(is_married) * 100),
    n_comorbid = sum(comorbidity),
    pct_comorbid = sprintf("%.1f%%", mean(comorbidity) * 100),
    n_white = sum(is_race_white),
    pct_white = sprintf("%.1f%%", mean(is_race_white) * 100),
    n_black = sum(is_race_black),
    pct_black = sprintf("%.1f%%", mean(is_race_black) * 100)
  ) %>%
  t() %>%
  as.data.frame()

colnames(table1_stratified) <- c("0 ε4 alleles", "1 ε4 allele", "2 ε4 alleles")

knitr::kable(table1_stratified,
             caption = "Table 1 (continued): Characteristics Stratified by APOE ε4 Status")
```

---

# 3. Model Fitting

This section fits the two-part model:
- **Part 1**: Logistic regression for dementia probability
- **Part 2**: Left-truncated right-censored (LTRC) model for timing among those with dementia

## Part 1: Dementia Probability (GLM)

```{r fit-part1}
cat("Fitting Part 1 model (logistic regression)...\n")
tic()

part_1_mod <- glm(
  got_dementia ~ t + years_education + is_female + is_married + comorbidity +
    as.factor(num_e4) + is_race_black + is_race_other + l,
  data = mod_dat,
  family = "binomial"
)

toc()

# Summary
summary(part_1_mod)

# Save model
write_rds(part_1_mod, fs::path(DATA_DIR, "nacc_pt_1_mod_sim.rds"))
cat("\n✓ Part 1 model saved to:", fs::path(DATA_DIR, "nacc_pt_1_mod_sim.rds"), "\n")
```

## Part 2: Timing of Dementia (LTRC)

NOTE: This takes ~5 minutes to run

```{r fit-part2}
cat("\nFitting Part 2 model (LTRC)...\n")
cat("⚠ NOTE: This takes ~5 minutes to run\n\n")

# Prepare data for Part 2 (dementia cases only)
dem_mod_dat <- mod_dat %>%
  filter(got_dementia == 1) %>%
  mutate(
    num_e4_1 = ifelse(num_e4 == 1, 1, 0),
    num_e4_2 = ifelse(num_e4 == 2, 1, 0)
  )

cat("Part 2 sample size:", nrow(dem_mod_dat), "(dementia cases only)\n\n")

tic()
part_2_mod <- ltrc(
  survival::Surv(mod_y, got_dementia) ~ t + years_education + is_female +
    is_married + comorbidity + num_e4_1 + num_e4_2 + is_race_black + is_race_other,
  data = dem_mod_dat,
  trunc_time = dem_mod_dat$mod_l,
  n_start = 10,
  int_knots = 1
)
toc()

# Get clean model
part_2_mod_clean <- get_clean_model(part_2_mod)

# Summary
cat("\nPart 2 Model Coefficients:\n")
print(part_2_mod_clean$parameters$beta)

cat("\nPart 2 Model Standard Errors:\n")
print(vcov(part_2_mod_clean) %>% diag() %>% sqrt())

# Save model
write_rds(part_2_mod_clean, fs::path(DATA_DIR, "nacc_pt_2_mod_sim.rds"))
cat("\n✓ Part 2 model saved to:", fs::path(DATA_DIR, "nacc_pt_2_mod_sim.rds"), "\n")
```

## Bootstrap Variance Estimation

Bootstrap resampling provides variance estimates for GHI predictions by repeatedly refitting the Part 2 model on resampled data. This is computationally intensive and best run on a server.

### Overview

The bootstrap procedure:
1. **Part 1 model**: Fit once on full data (not resampled)
2. **Part 2 model**: Resample dementia cases with replacement and refit
3. **Predictions**: Generate GHI estimates across full covariate grid for each bootstrap sample
4. **Variance**: Calculate empirical variance of GHI estimates across bootstrap samples

**For this notebook**: We skip bootstrap computation and proceed with model-based standard errors only. Bootstrap results can be loaded if previously computed.

### Check for Pre-computed Bootstrap Results

```{r bootstrap-check}
# Check if bootstrap results exist
bs_dir <- "inst/sim_res/NACC_mod_bs_with_race"
if (fs::dir_exists(bs_dir) && length(fs::dir_ls(bs_dir)) > 0) {
  cat("✓ Bootstrap results found. Loading...\n")

  bs_res <- data.frame()
  for (f in fs::dir_ls(bs_dir)) {
    tmp <- read_csv(f, show_col_types = FALSE)
    bs_res <- bs_res %>% bind_rows(tmp)
  }

  cat("✓ Loaded", nrow(bs_res), "bootstrap samples\n")
  cat("  Unique grid points:", n_distinct(bs_res %>% select(t, educ, is_female, is_married,
                                                            has_comorbidity, e4_alleles_1,
                                                            e4_alleles_2, is_race_black,
                                                            is_race_other, l)), "\n")
  BOOTSTRAP_AVAILABLE <- TRUE
} else {
  cat("ℹ No bootstrap results found.\n")
  cat("  Proceeding without bootstrap variance estimates.\n")
  cat("  Model-based standard errors will be used instead.\n\n")
  cat("To compute bootstrap estimates, see server instructions below.\n")
  BOOTSTRAP_AVAILABLE <- FALSE
}
```

### Server-Based Bootstrap (Production Method)

The production bootstrap used the following specification:
- **Bootstrap iterations per file**: 6
- **Total files**: 50 (yielding 300 total bootstrap samples)
- **Resampling**: Dementia cases only (Part 2 model)
- **Prediction grid**: 2,688 covariate combinations

```markdown
# Large-Scale Bootstrap on Server

## Setup

1. Prepare data file:
   - Ensure `NACC_mod_dat.csv` is in working directory
   - Should contain the full modeling dataset

2. Upload scripts to server:
   - `06_NACC_bootstrap_on_server.R` (main bootstrap code)
   - `07_run_NACC_bootstrap_parallel.sh` (parallel execution script)

3. Create output directory:
   ```bash
   mkdir -p NACC_mod_bs
   ```

4. Review prediction grid in `06_NACC_bootstrap_on_server.R`:
   ```r
   pred_grid <- list(
     t = 5:25,                    # Age at death - 65 (21 values)
     educ = 16,                   # Education fixed at 16 years
     is_female = c(0, 1),         # Both sexes
     is_married = c(0, 1),        # Married status
     has_comorbidity = c(0, 1),   # Comorbidity
     e4_alleles_1 = c(0, 1),      # One ε4 allele
     e4_alleles_2 = c(0, 1),      # Two ε4 alleles
     is_race_black = c(0, 1),     # Black race
     is_race_other = 0,           # Other race fixed at 0
     l = 1                        # Entry age - 65 fixed at 1
   ) %>% expand.grid()
   ```
   Total: 21 × 1 × 2 × 2 × 2 × 2 × 2 × 2 × 1 = **2,688 grid points**

## Running

### Option A: Bash Script (Background Jobs)

```bash
cd inst/paper_code
bash 07_run_NACC_bootstrap_parallel.sh
wait  # Wait for all jobs to complete
```

### Option B: Manual Parallel Execution

```bash
# Run 50 files in parallel (FILE_NUM = 1 to 50)
for i in {1..50}; do
  Rscript 06_NACC_bootstrap_on_server.R $i > bs_out_$i.txt 2>&1 &
done
wait
```

## Expected Runtime

- **Per file** (6 bootstrap iterations × 2,688 predictions): 30-60 minutes
- **Total (50 parallel jobs)**: 30-60 minutes wall time
- **Total (sequential)**: 25-50 hours

## Output

- 50 `.csv` files in `NACC_mod_bs/`
- Each file: `file_{FILE_NUM}.csv`
- Each contains 6 bootstrap iterations × 2,688 grid points = 16,128 rows
- Total: 50 × 16,128 = 806,400 bootstrap predictions

## Post-Processing

After bootstrap completes:

1. **Combine files**: All files can be loaded together (as done in bootstrap-check chunk above)

2. **Calculate variance**: Group by grid point and compute empirical variance
   ```r
   bs_variance <- bs_res %>%
     group_by(t, educ, is_female, is_married, has_comorbidity,
              e4_alleles_1, e4_alleles_2, is_race_black, is_race_other, l) %>%
     summarize(
       mean_ghi = mean(ghi_estimate),
       var_ghi = var(ghi_estimate),
       sd_ghi = sd(ghi_estimate),
       .groups = "drop"
     )
   ```

3. **Use in figures**: Bootstrap SE provides confidence bands for GHI plots
```

---

# 4. Model Analysis

This section generates **Table 2** (coefficient estimates with confidence intervals) and **Figures 1-2** from the paper.

## Table 2: Model Coefficients

```{r table2}
# Load fitted models if not already in memory
if (!exists("part_1_mod")) {
  part_1_mod <- read_rds(fs::path(DATA_DIR, "nacc_pt_1_mod.rds"))
}
if (!exists("part_2_mod_clean")) {
  part_2_mod_clean <- read_rds(fs::path(DATA_DIR, "nacc_pt_2_mod.rds"))
}

# Part 2 coefficients
omegas <- part_2_mod_clean$parameters$beta
sds_omega <- vcov(part_2_mod_clean) %>% diag() %>% sqrt()

# Part 1 coefficients (excluding intercept)
betas <- part_1_mod$coefficients[-1]
sds_beta <- vcov(part_1_mod) %>% diag() %>% sqrt() %>% `[`(-1)

# Create coefficient table
coef_table <- data.frame(
  Covariate = c("t (time since age 65)", "Years of education", "Female",
                "Married", "≥1 Comorbidity", "1 ε4 allele", "2 ε4 alleles",
                "Black race", "Other race", "l (left truncation)"),
  Part1_Est = round(betas, 3),
  Part1_CI = paste0("(", round(betas - 1.96 * sds_beta, 3), ", ",
                    round(betas + 1.96 * sds_beta, 3), ")"),
  Part1_P = format.pval(2 * pnorm(-abs(betas / sds_beta)), digits = 3, eps = 0.001),
  Part2_Est = c(round(omegas, 3), "—"),
  Part2_CI = c(paste0("(", round(omegas - 1.96 * sds_omega, 3), ", ",
                      round(omegas + 1.96 * sds_omega, 3), ")"), "—"),
  Part2_P = c(format.pval(2 * pnorm(-abs(omegas / sds_omega)), digits = 3, eps = 0.001), "—")
)

knitr::kable(coef_table,
             caption = "Table 2: Coefficient Estimates for Two-Part Model",
             col.names = c("Covariate", "Estimate", "95% CI", "P-value",
                          "Estimate", "95% CI", "P-value"),
             align = c("l", "r", "l", "r", "r", "l", "r"))
```

## Prediction Function

Define the function to compute GHI predictions:

```{r prediction-function}
# Update dem_mod_dat with full model data
mod_dat <- mod_dat %>%
  mutate(
    num_e4_1 = ifelse(num_e4 == 1, 1, 0),
    num_e4_2 = ifelse(num_e4 == 2, 1, 0)
  )
dem_mod_dat <- mod_dat %>% filter(got_dementia == 1)

predict_ghi <- function(mod1, mod2, t, educ, is_female, is_married,
                       has_comorbidity, e4_alleles_1, e4_alleles_2,
                       is_race_black, is_race_other, l) {

  # Part 1: Probability of dementia
  lp_mu <- sum(mod1$coefficients[-1] *
                 c(t, educ, is_female, is_married, has_comorbidity,
                   e4_alleles_1, e4_alleles_2, is_race_black, is_race_other, l))
  mu <- exp(lp_mu) / (1 + exp(lp_mu))

  var_mu <- (mu * (1 - mu))^2 *
    c(1, t, educ, is_female, is_married, has_comorbidity,
      e4_alleles_1, e4_alleles_2, is_race_black, is_race_other, l) %*%
    vcov(mod1) %*%
    c(1, t, educ, is_female, is_married, has_comorbidity,
      e4_alleles_1, e4_alleles_2, is_race_black, is_race_other, l)

  # Part 2: Expected timing given dementia
  lp_y <- sum(mod2$parameters$beta *
                c(t, educ, is_female, is_married, has_comorbidity,
                  e4_alleles_1, e4_alleles_2, is_race_black, is_race_other))

  sd_lp_y <- c(t, educ, is_female, is_married, has_comorbidity,
               e4_alleles_1, e4_alleles_2, is_race_black, is_race_other) %*%
    vcov(mod2) %*%
    c(t, educ, is_female, is_married, has_comorbidity,
      e4_alleles_1, e4_alleles_2, is_race_black, is_race_other)

  # Kaplan-Meier integration for E[Y | dementia]
  mod_l <- log((l/t) / (1 - (l/t)))
  lower_bound <- mod_l - lp_y

  tau <- dem_mod_dat %>%
    mutate(l_resid_scale = mod_l - lp_y) %>%
    pull(l_resid_scale)

  km_df <- data.frame(
    epsilon = mod2$data$residuals,
    tau = tau,
    delta = 1
  ) %>%
    mutate(
      g_of_e = exp(lp_y + epsilon) / (1 + exp(lp_y + epsilon)),
      g_of_t = exp(lp_y + tau) / (1 + exp(lp_y + tau))
    ) %>%
    filter(epsilon > lower_bound, g_of_e > g_of_t)

  if (nrow(km_df) < 3) return(data.frame())

  km_fit <- tryCatch({
    survfit(Surv(time = g_of_t, time2 = g_of_e, event = delta) ~ 1, data = km_df)
  }, error = function(e) return(NULL))

  if (is.null(km_fit)) return(data.frame())

  km_times <- c(0, km_fit$time)
  km_surv <- c(1, km_fit$surv)
  km_estimate <- sum(km_surv[-length(km_surv)] * diff(km_times))

  # GHI = E[min(T, S)] = (1-mu) + mu * E[Y | dementia]
  ghi_estimate <- (1 - mu) + mu * km_estimate

  data.frame(
    t = t, educ = educ, is_female = is_female, is_married = is_married,
    has_comorbidity = has_comorbidity, e4_alleles_1 = e4_alleles_1,
    e4_alleles_2 = e4_alleles_2, is_race_black = is_race_black,
    is_race_other = is_race_other, l = l,
    lp_mu = lp_mu, mu = mu, var_mu = var_mu,
    lp_y = lp_y, sd_lp_y = sd_lp_y,
    expected_y = km_estimate,
    ghi_estimate = ghi_estimate
  )
}
```

## Generate Predictions

```{r generate-predictions}
cat("Generating predictions across covariate grid...\n")

pred_grid <- expand.grid(
  t = c(0.5, 1:25),
  educ = 16,
  is_female = c(0, 1),
  is_married = c(0, 1),
  has_comorbidity = c(0, 1),
  e4_alleles_1 = c(0, 1),
  e4_alleles_2 = c(0, 1),
  is_race_black = c(0, 1),
  is_race_other = c(0,1),
  l = 1
)

# Remove invalid combinations (can't have both e4 alleles)
pred_grid <- pred_grid %>%
  filter(!(e4_alleles_1 == 1 & e4_alleles_2 == 1))

cat("Prediction grid size:", nrow(pred_grid), "combinations\n")
cat("Computing predictions (this may take 1-2 minutes)...\n\n")

tic()
res <- pred_grid %>%
  pmap_dfr(function(...) {
    args <- list(...)
    predict_ghi(
      part_1_mod, part_2_mod_clean,
      t = args$t, educ = args$educ, is_female = args$is_female,
      is_married = args$is_married, has_comorbidity = args$has_comorbidity,
      e4_alleles_1 = args$e4_alleles_1, e4_alleles_2 = args$e4_alleles_2,
      is_race_black = args$is_race_black, is_race_other = args$is_race_other,
      l = args$l
    )
  })
toc()

cat("\n✓ Generated", nrow(res), "predictions\n")
```

## Figures 1 and 2: GHI by Age, Race, Sex, Marital Status, Comorbidity, and APOE Status


```{r}
fig1 <- res %>%
  mutate(
    Race = case_when(
      is_race_black == 1 ~ "Race: Black",
      is_race_other == 1 ~ "Race: Other",
      TRUE ~ "Race: White"
    ),
    Race = factor(Race, levels = c("Race: White", "Race: Black", "Race: Other")),
    married_str = ifelse(is_married == 1, "Married", "Unmarried"),
    sex_str = ifelse(is_female == 1, "Female", "Male"),
    comorbidity_str = ifelse(has_comorbidity == 1, "with Comorbidity", "without Comorbidity"),
    Group = paste0(married_str, " ", sex_str, "\n", comorbidity_str)
  ) %>%
  filter(is_female == 0) %>%
  filter(educ == 16) %>%
  filter(!(e4_alleles_2 == 1 & e4_alleles_1 == 1)) %>%
  filter(!(is_race_black == 1 & is_race_other == 1)) %>%
  mutate(
    num_e4 = case_when(
      e4_alleles_2 == 1 ~ 2,
      e4_alleles_1 == 1 ~ 1,
      TRUE ~ 0
    ),
    Comorbidity = ifelse(has_comorbidity == 1, "Yes", "No"),
    Sex = ifelse(is_female == 1, "Female", "Male")#,
    # lower = ghi_estimate - 2 * sd_ghi,
    # upper = ghi_estimate + 2 * sd_ghi
  ) %>%
  ggplot() +
  aes(x = t + 65, y = ghi_estimate, color = as.factor(num_e4), fill = as.factor(num_e4)) +
  # {if(all(!is.na(res_w_var$sd_ghi))) geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, color = NA)} +
  geom_line(aes(lty = as.factor(num_e4)), size = 1) +
  facet_grid(Group ~ Race, labeller = "label_value") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times"),
    strip.background = element_blank(),
    strip.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(x = "Age at Death", y = "Predicted GHI", color = "Number of e4 Alleles", fill = "Number of e4 Alleles", lty = "Number of e4 Alleles") +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  xlim(c(65, 90))

print(fig1)
      # Save figure
ggsave(fs::path(RESULTS_DIR, "figure1.png"),
       fig1, width = 12, height = 7, dpi = 300)

fig2 <- res %>%
  mutate(
    Race = case_when(
      is_race_black == 1 ~ "Race: Black",
      is_race_other == 1 ~ "Race: Other",
      TRUE ~ "Race: White"
    ),
    Race = factor(Race, levels = c("Race: White", "Race: Black", "Race: Other")),
    married_str = ifelse(is_married == 1, "Married", "Unmarried"),
    sex_str = ifelse(is_female == 1, "Female", "Male"),
    comorbidity_str = ifelse(has_comorbidity == 1, "with Comorbidity", "without Comorbidity"),
    Group = paste0(married_str, " ", sex_str, "\n", comorbidity_str)
  ) %>%
  filter(is_female == 1) %>%
  filter(educ == 16) %>%
  filter(!(e4_alleles_2 == 1 & e4_alleles_1 == 1)) %>%
  filter(!(is_race_black == 1 & is_race_other == 1)) %>%
  mutate(
    num_e4 = case_when(
      e4_alleles_2 == 1 ~ 2,
      e4_alleles_1 == 1 ~ 1,
      TRUE ~ 0
    ),
    Comorbidity = ifelse(has_comorbidity == 1, "Yes", "No"),
    Sex = ifelse(is_female == 1, "Female", "Male")#,
    # lower = ghi_estimate - 2 * sd_ghi,
    # upper = ghi_estimate + 2 * sd_ghi
  ) %>%
  ggplot() +
  aes(x = t + 65, y = ghi_estimate, color = as.factor(num_e4), fill = as.factor(num_e4)) +
  # {if(all(!is.na(res_w_var$sd_ghi))) geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, color = NA)} +
  geom_line(aes(lty = as.factor(num_e4)), size = 1) +
  facet_grid(Group ~ Race, labeller = "label_value") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times"),
    strip.background = element_blank(),
    strip.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(x = "Age at Death", y = "Predicted GHI", color = "Number of e4 Alleles", fill = "Number of e4 Alleles", lty = "Number of e4 Alleles") +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  xlim(c(65, 90))

print(fig2)
      # Save figure
ggsave(fs::path(RESULTS_DIR, "figure2.png"),
       fig2, width = 12, height = 7, dpi = 300)
```

---

# 5. Model Checking

Comprehensive diagnostics for both parts of the model.

## Part 1 Diagnostics {.tabset}

### Surrogate Residuals

```{r part1-surrogate-residuals}
cat("Computing surrogate residuals for Part 1 model...\n")

# Refit Part 1 with y = TRUE to access response
part_1_mod_check <- glm(
  got_dementia ~ t + years_education + is_female + is_married + comorbidity +
    as.factor(num_e4) + is_race_black + is_race_other + l,
  data = mod_dat,
  family = "binomial",
  y = TRUE
)

# Surrogate residuals function
surrogate_residuals_logistic <- function(model, nsim = 1, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  y <- model$y
  if (is.null(y)) stop("Model must be fit with y = TRUE")

  p <- fitted(model)
  n <- length(y)
  res <- matrix(NA_real_, nrow = n, ncol = nsim)

  for (s in seq_len(nsim)) {
    U <- numeric(n)
    for (i in seq_len(n)) {
      if (y[i] == 0) {
        U[i] <- runif(1, min = 0, max = 1 - p[i])
      } else {
        U[i] <- runif(1, min = 1 - p[i], max = 1)
      }
    }
    res[, s] <- qnorm(U)
  }

  if (nsim == 1) drop(res) else res
}

set.seed(26339)
r <- surrogate_residuals_logistic(part_1_mod_check, seed = 26339)
R <- surrogate_residuals_logistic(part_1_mod_check, nsim = 100, seed = 26339)

cat("✓ Computed surrogate residuals\n")
```

### QQ Plot

```{r part1-qq, fig.width=8, fig.height=6}
# QQ plot envelope
qq_data <- apply(R, 2, sort)
env_bottom <- apply(qq_data, 1, min)
env_top <- apply(qq_data, 1, max)

qq_plot <- ggplot() +
  aes(sample = r) +
  geom_ribbon(
    aes(x = qnorm(ppoints(nrow(R))), ymin = env_bottom, ymax = env_top),
    fill = "lightblue",
    alpha = 0.5
  ) +
  geom_qq(alpha = 0.3, size = 1) +
  geom_qq_line(color = "red", lty = "dashed", size = 1) +
  labs(
    title = "Part 1: QQ Plot of Surrogate Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_bw(base_size = 14)

print(qq_plot)
```

### Hosmer-Lemeshow Test

```{r part1-hosmer-lemeshow}
# Hosmer-Lemeshow goodness of fit test
hl_test <- hoslem.test(part_1_mod_check$y, fitted(part_1_mod_check), g = 10)

cat("Hosmer-Lemeshow Test:\n")
cat("  Chi-squared:", round(hl_test$statistic, 3), "\n")
cat("  df:", hl_test$parameter, "\n")
cat("  p-value:", format.pval(hl_test$p.value, digits = 3), "\n\n")

if (hl_test$p.value > 0.05) {
  cat("✓ Good fit (fail to reject null hypothesis of good calibration)\n")
} else {
  cat("⚠ Potential calibration issues (reject null hypothesis)\n")
}
```

### Calibration Plot

```{r part1-calibration, fig.width=8, fig.height=6}
hl <- hoslem.test(
  x = mod_dat$got_dementia,
  y = fitted(part_1_mod),
  g = 10
)

hl_data <- data.frame(
  observed = hl$observed[, 2],
  expected = hl$expected[, 2],
  group = seq_len(nrow(hl$observed))
)

calib <- mod_dat %>%
  mutate(
    p_hat = fitted(part_1_mod),
    group = ntile(p_hat, 10)
  ) %>%
  group_by(group) %>%
  summarize(
    mean_p = mean(p_hat),
    obs_rate = mean(got_dementia),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    se = sqrt(obs_rate * (1 - obs_rate) / n),
    lower = obs_rate - 1.96 * se,
    upper = obs_rate + 1.96 * se
  )

cal_plot <- calib %>%
  ggplot() +
  geom_point(
    aes(x = mean_p, y = obs_rate),
    size = 2
  ) +
  geom_segment(
    aes(x = mean_p, xend = mean_p, y = lower, yend = upper)
  ) +
  geom_segment(
    aes(x = mean_p - 0.005, xend = mean_p + 0.005, y = lower, yend = lower)
  ) +
  geom_segment(
    aes(x = mean_p - 0.005, xend = mean_p + 0.005, y = upper, yend = upper)
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", lty = "dashed") +
  xlab("Mean Predicted Probability") +
  ylab("Observed Event Rate") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times"),
    strip.background = element_blank(),
    strip.text = element_text(size = 15),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 18)
  )

print(cal_plot)
```

### VIF (Multicollinearity)

```{r part1-vif}
# Variance Inflation Factors
vif_values <- car::vif(part_1_mod_check)

cat("Variance Inflation Factors (VIF):\n")
print(round(vif_values, 2))

max_vif <- max(vif_values)
cat("\nMax VIF:", round(max_vif, 2), "\n")

if (max_vif < 5) {
  cat("✓ No multicollinearity concerns (all VIF < 5)\n")
} else if (max_vif < 10) {
  cat("⚠ Moderate multicollinearity (some VIF between 5-10)\n")
} else {
  cat("⚠ Severe multicollinearity (some VIF > 10)\n")
}
```

### Cook's Distance

```{r part1-cooks, fig.width=10, fig.height=6}
# Cook's distance
cooks_d <- cooks.distance(part_1_mod_check)

# Cutoff: 4 / n
cutoff <- 4 / nrow(mod_dat)
influential <- which(cooks_d > cutoff)

cat("Cook's Distance Analysis:\n")
cat("  Cutoff (4/n):", round(cutoff, 6), "\n")
cat("  Influential points:", length(influential), "of", nrow(mod_dat), "\n")

# Plot
cooks_plot <- ggplot(data.frame(index = 1:length(cooks_d), cooks = cooks_d),
                     aes(x = index, y = cooks)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = cutoff, color = "red", lty = "dashed") +
  labs(
    title = "Part 1: Cook's Distance",
    subtitle = paste0(length(influential), " influential observations (red line = 4/n cutoff)"),
    x = "Observation Index",
    y = "Cook's Distance"
  ) +
  theme_bw(base_size = 14)

print(cooks_plot)

# Refit without influential points
if (length(influential) > 0 && length(influential) < nrow(mod_dat) * 0.1) {
  cat("\nRefitting model without influential points...\n")
  mod_dat_no_influential <- mod_dat[-influential, ]

  part_1_mod_robust <- glm(
    got_dementia ~ t + years_education + is_female + is_married + comorbidity +
      as.factor(num_e4) + is_race_black + is_race_other + l,
    data = mod_dat_no_influential,
    family = "binomial"
  )

  # Compare coefficients
  coef_comparison <- data.frame(
    Original = coef(part_1_mod_check),
    Robust = coef(part_1_mod_robust),
    Difference = coef(part_1_mod_robust) - coef(part_1_mod_check),
    Pct_Change = (coef(part_1_mod_robust) - coef(part_1_mod_check)) / coef(part_1_mod_check) * 100
  )

  cat("\nCoefficient comparison (original vs. robust):\n")
  print(round(coef_comparison, 4))

  max_pct_change <- max(abs(coef_comparison$Pct_Change), na.rm = TRUE)
  if (max_pct_change < 10) {
    cat("\n✓ Robust model similar (max change < 10%)\n")
  } else {
    cat("\n⚠ Substantial changes detected (max change > 10%)\n")
  }
}
```

Note that while there are some substantial changes, they occur in groups that had a small number of individuals to begin with. Further, the key here is to assess if outliers are impacting our main conclusions for the APOE covariates, which they are not since the robust estimates are in the same direction, and actually have a larger magnitude when fit.

### Linearity Check

```{r part1-linearity, fig.width=12, fig.height=8}
# Check linearity assumption for continuous predictors
eta <- predict(part_1_mod_check, type = "link")

linearity_plots <- list()

# t
p1 <- ggplot(data.frame(t = mod_dat$t, r = r), aes(x = t, y = r)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  geom_hline(yintercept = 0, color = "red", lty = "dashed") +
  labs(title = "Residuals vs. t", x = "t", y = "Surrogate Residual") +
  theme_bw()

# Education
p2 <- ggplot(data.frame(educ = mod_dat$years_education, r = r),
             aes(x = educ, y = r)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  geom_hline(yintercept = 0, color = "red", lty = "dashed") +
  labs(title = "Residuals vs. Education", x = "Years of Education",
       y = "Surrogate Residual") +
  theme_bw()

# l
p3 <- ggplot(data.frame(l = mod_dat$l, r = r), aes(x = l, y = r)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  geom_hline(yintercept = 0, color = "red", lty = "dashed") +
  labs(title = "Residuals vs. l", x = "l", y = "Surrogate Residual") +
  theme_bw()

# Linear predictor
p4 <- ggplot(data.frame(eta = eta, r = r), aes(x = eta, y = r)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  geom_hline(yintercept = 0, color = "red", lty = "dashed") +
  labs(title = "Residuals vs. Linear Predictor", x = "Linear Predictor",
       y = "Surrogate Residual") +
  theme_bw()

# Combine plots
print(p1)
print(p2)
print(p3)
print(p4)
```

### Likelihood Ratio Test (Splines)

```{r part1-lrt-splines}
cat("Testing spline alternatives for continuous predictors...\n\n")

# Test splines for t
part_1_mod_t_spline <- glm(
  got_dementia ~ ns(t, df = 3) + years_education + is_female + is_married +
    comorbidity + as.factor(num_e4) + is_race_black + is_race_other + l,
  data = mod_dat,
  family = "binomial"
)

lrt_t <- anova(part_1_mod_check, part_1_mod_t_spline, test = "Chisq")
cat("LRT for t (linear vs. spline):\n")
print(lrt_t)

# Test splines for education
part_1_mod_educ_spline <- glm(
  got_dementia ~ t + ns(years_education, df = 3) + is_female + is_married +
    comorbidity + as.factor(num_e4) + is_race_black + is_race_other + l,
  data = mod_dat,
  family = "binomial"
)

lrt_educ <- anova(part_1_mod_check, part_1_mod_educ_spline, test = "Chisq")
cat("\nLRT for education (linear vs. spline):\n")
print(lrt_educ)

# Test splines for l
part_1_mod_l_spline <- glm(
  got_dementia ~ t + years_education + is_female + is_married + comorbidity +
    as.factor(num_e4) + is_race_black + is_race_other + ns(l, df = 3),
  data = mod_dat,
  family = "binomial"
)

lrt_l <- anova(part_1_mod_check, part_1_mod_l_spline, test = "Chisq")
cat("\nLRT for l (linear vs. spline):\n")
print(lrt_l)

cat("\n")
if (lrt_t$`Pr(>Chi)`[2] > 0.05 && lrt_educ$`Pr(>Chi)`[2] > 0.05 &&
    lrt_l$`Pr(>Chi)`[2] > 0.05) {
  cat("✓ Linear specifications adequate for all continuous predictors\n")
} else {
  cat("⚠ Some non-linear relationships detected - consider splines\n")
}
```

## Part 2 Diagnostics {.tabset}

### Partial Martingale Residual Plots

```{r}
X <- model.matrix(~ . - 1, data = as.data.frame(part_2_mod_clean$data$predictors))
beta <- part_2_mod_clean$parameters$beta
resid <- as.numeric(part_2_mod_clean$data$residuals)

partial_resid_plot <- function(j, var_name) {

  if (var_name == "Age at Death") add_val <- 65
  else add_val <- 0
  pr <- resid + X[, j] * beta[j]

  plot_dat <- data.frame(
    x = X[, j],
    pr = pr
  )

  ggplot(plot_dat, aes(x = x + add_val, y = pr)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "loess", se = FALSE, color = "blue") +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red") +
    labs(
      x = var_name,
      y = "Partial residual"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      text = element_text(family = "Times"),
      strip.background = element_blank(),
      strip.text = element_text(size = 15),
      axis.text = element_text(size = 15),
      plot.title = element_text(hjust = 0.5, size = 20),
      axis.title = element_text(size = 18)
    )
}

# Example
partial_resid_plot(j = 1, var_name = "Age at Death")
partial_resid_plot(j = 2, var_name = "Years of Education")
```


### Cumulative Martingale Processes

```{r part2-cumulative-mart, fig.width=10, fig.height=6}
cum_mart_resid_plot <- function(mart, x, x_label = "Age at Death") {
  n <- length(mart)

  # jitter if the variable is discrete
  if (length(x) != length(unique(x))) {
    x <- x + rnorm(length(x), 0, 0.1)
  }

  ord <- order(x)
  x_ord <- x[ord]
  m_ord <- mart[ord]

  R_obs <- cumsum(m_ord) / sqrt(n)

  B <- 1000
  R_sim <- matrix(NA_real_, nrow = n, ncol = B)

  for (b in seq_len(B)) {
    G <- rnorm(n)
    M_star <- m_ord * G
    R_sim[, b] <- cumsum(M_star) / sqrt(n)
  }

  lower <- apply(R_sim, 1, quantile, probs = 0.025)
  upper <- apply(R_sim, 1, quantile, probs = 0.975)

  if (x_label == "Age at Death") plot_x <- x_ord + 65
  else plot_x <- x_ord

  p <- ggplot() +
    geom_ribbon(
      aes(x = plot_x, ymin = lower, ymax = upper),
      fill = "blue",
      alpha = 0.25
    ) +
    geom_line(
      aes(x = plot_x, y = R_obs)
    ) +
    geom_hline(yintercept = 0, color = "red", lty = "dashed") +
    xlab(x_label) +
    ylab("Cumulative Martingale Residual") +
    theme_bw() +
    theme(
      legend.position = "bottom",
      text = element_text(family = "Times"),
      strip.background = element_blank(),
      strip.text = element_text(size = 15),
      axis.text = element_text(size = 15),
      axis.title = element_text(size = 18)
    )

  return(p)
}

martingale_resids <- part_2_mod_clean$data$martingale_residuals
cum_mart_resid_plot(martingale_resids, dem_mod_dat$t, x_label = "Age at Death")
cum_mart_resid_plot(martingale_resids, dem_mod_dat$years_education, x_label = "Years of Education")

```

### Influence Diagnostics

NOTE: This will take about 5 minutes to run

```{r part2-influence}
# Identify high-leverage points and outliers
# Note: This is approximate - full dfbeta analysis requires custom implementation

cat("Part 2 Influence Diagnostics:\n\n")

# Leverage (hat values) - approximated using residuals
mart_resid <- part_2_mod_clean$data$martingale_residuals
leverage_approx <- abs(mart_resid) / sd(mart_resid)
high_leverage <- which(leverage_approx > 2.5)

cat("High leverage points (|standardized residual| > 2.5):", length(high_leverage), "\n")
cat("Proportion:", round(length(high_leverage) / length(mart_resid) * 100, 2), "%\n")

if (length(high_leverage) > 0 && length(high_leverage) < nrow(dem_mod_dat) * 0.1) {
  cat("\nRefitting model without high-leverage points...\n")

  dem_mod_dat_robust <- dem_mod_dat[-high_leverage, ]

  tic()
  part_2_mod_robust <- ltrc(
    survival::Surv(mod_y, got_dementia) ~ t + years_education + is_female +
      is_married + comorbidity + num_e4_1 + num_e4_2 + is_race_black + is_race_other,
    data = dem_mod_dat_robust,
    trunc_time = dem_mod_dat_robust$mod_l,
    n_start = 10,
    int_knots = 1
  )
  toc()

  part_2_mod_robust_clean <- get_clean_model(part_2_mod_robust)

  # Compare coefficients
  coef_comparison_p2 <- data.frame(
    Original = part_2_mod_clean$parameters$beta,
    Robust = part_2_mod_robust_clean$parameters$beta,
    Difference = part_2_mod_robust_clean$parameters$beta - part_2_mod_clean$parameters$beta,
    Pct_Change = (part_2_mod_robust_clean$parameters$beta - part_2_mod_clean$parameters$beta) /
                 part_2_mod_clean$parameters$beta * 100
  )

  cat("\nCoefficient comparison (original vs. robust):\n")
  print(round(coef_comparison_p2, 4))

  max_pct_change <- max(abs(coef_comparison_p2$Pct_Change))
  if (max_pct_change < 10) {
    cat("\n✓ Robust model similar (max change < 10%)\n")
  } else {
    cat("\n⚠ Substantial changes detected (max change > 10%)\n")
  }
} else {
  cat("\n⚠ Too many or too few high-leverage points for robust comparison\n")
}
```

These results are similar to what we had in this same analysis for the part 1 model. Some large changes but not relative to the standard error of the estimate. And the conclusion remains the same.

### Comparison to Parametric Model

```{r part2-parametric, eval=FALSE}
cat("Comparing semiparametric LTRC to parametric truncated regression...\n")

resids <- as.numeric(part_2_mod_clean$data$residuals)
# need to subtract the linear predictor to get the truncation time on the same scale as the residuals
trunc_points <- dem_mod_dat$mod_l - part_2_mod_clean$data$fitted_response
sum(trunc_points > resids) # sanity check

# Estimate parameters assuming truncated normal (MLE under truncation)
# Using method of moments as starting point, then refining
resid_mean_init <- mean(resids)
resid_sd_init <- sd(resids)

# Fit truncated normal via maximum likelihood
neg_loglik <- function(par, t, l) {
  mu <- par[1]
  sigma <- exp(par[2])

  z_t <- (t - mu) / sigma
  z_l <- (l - mu) / sigma

  ll <- sum(
    dnorm(z_t, log = TRUE) - log(sigma) -
      pnorm(z_l, lower.tail = FALSE, log.p = TRUE)
  )

  return(-ll)
}

# Optimize
fit_trunc <- optim(
  par = c(resid_mean_init, log(resid_sd_init)),
  fn = neg_loglik,
  t = resids,
  l = trunc_points,
  method = "BFGS"
)

mu_hat <- fit_trunc$par[1]
sigma_hat <- exp(fit_trunc$par[2])

cat("Fitted truncated normal parameters:\n")
cat("  mu =", round(mu_hat, 4), "\n")
cat("  sigma =", round(sigma_hat, 4), "\n\n")

# Transform residuals to uniform via truncated normal CDF, then to standard normal
# If residuals truly follow TruncNorm(mu, sigma, a=l_i), then z should be N(0,1)
u <- (pnorm((resids - mu_hat) / sigma_hat) - pnorm((trunc_points - mu_hat) / sigma_hat)) /
  (1 - pnorm((trunc_points - mu_hat) / sigma_hat))
z <- qnorm(u)

# Remove any infinite/NA values
z_clean <- z[is.finite(z)]

# Create QQ plot data for ggplot
n <- length(z_clean)
theoretical_quantiles <- qnorm(ppoints(n))
observed_sorted <- sort(z_clean)

qq_df <- data.frame(
  theoretical = theoretical_quantiles,
  observed = observed_sorted
)

# QQ Plot with ggplot
qq_trunc_plot <- ggplot(qq_df) +
  geom_point(aes(y = theoretical, x = observed), alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red", lty = "dashed", linewidth = 1) +
  labs(
    title = "Part 2: QQ Plot of Residuals vs. Truncated Normal",
    subtitle = "If residuals follow truncated normal, points should lie on the line",
    y = "Theoretical Quantiles (Standard Normal)",
    x = "Transformed Residuals"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times"),
    strip.background = element_blank(),
    strip.text = element_text(size = 15),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 18),
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 12)
  )

print(qq_trunc_plot)
```

---

# 6. Sensitivity Analyses

## 6.1 ADRC Sub-Analysis {.tabset}

### Included vs. Excluded Subjects

```{r adrc-comparison}
cat("ADRC Comparison Analysis\n")
cat("========================\n\n")

cat("⚠ Note: With simulated data, subject-level characteristics are artificial\n")
cat("  For real data, compare demographics of included vs. excluded subjects\n\n")

# With real data, would compare:
# - Age distributions
# - Sex ratios
# - Education levels
# - Dementia rates
# - Between those in the analysis vs. those excluded

cat("Analysis to perform with real NACC data:\n")
cat("  1. Identify exclusion reasons (missing data, age < 65, etc.)\n")
cat("  2. Compare demographics: age, sex, education, race, APOE\n")
cat("  3. Assess potential selection bias\n")
cat("  4. Consider sensitivity to exclusion criteria\n")
```

### Random Effects Models

Fit mixed-effects models with random intercepts and slopes by ADRC:

```{r adrc-random-effects}
cat("Fitting random effects models by ADRC...\n\n")

# Load additional packages
library(statmod)
library(truncnorm)
library(TMB)

# Part 1: Random intercept logistic regression
cat("Part 1: Random intercept logistic model\n")
cat("---------------------------------------\n")
tic()
part_1_re <- glmer(
  got_dementia ~ t + years_education + is_female + is_married + comorbidity +
    as.factor(num_e4) + is_race_black + is_race_other + l + (1 | adc),
  data = mod_dat,
  family = binomial
)
# This sometimes errors out, just try it a couple of times
toc()

summary(part_1_re)

# Extract fixed effects and random effects variance
fe_p1 <- fixef(part_1_re)
se_p1 <- sqrt(diag(vcov(part_1_re)))
re_var_p1 <- as.data.frame(VarCorr(part_1_re))

cat("\n\nPart 1 Random Effects Variance:\n")
print(re_var_p1)

# Part 2: Truncated regression with random effects
cat("\n\nPart 2: Truncated regression with random effects\n")
cat("------------------------------------------------\n")
cat("⚠ Note: This uses TMB for random slopes - may take 2-3 minutes\n\n")

# First, fit basic truncated regression (no random effects) as baseline
library(truncreg)
fit_basic <- truncreg(
  mod_y ~ t + years_education + is_female + is_married + comorbidity +
    num_e4_1 + num_e4_2 + is_race_black + is_race_other,
  point = dem_mod_dat$mod_l,
  data = dem_mod_dat,
  direction = "left"
)

cat("Basic truncated regression (no random effects):\n")
print(summary(fit_basic))

# Now fit truncated LMM with random intercept and slopes using TMB
cat("\n\nFitting truncated LMM with random effects using TMB...\n")

# Compile TMB model
cpp_file <- "model/truncated_lmm.cpp"
if (file.exists(cpp_file)) {
  cat("Compiling TMB model...\n")
  TMB::compile(cpp_file)
  dyn.load(dynlib("model/truncated_lmm"))

  # Construct fixed and random effects design matrices
  X <- model.matrix(~ t + years_education + is_female + is_married +
                      comorbidity + num_e4_1 + num_e4_2 +
                      is_race_black + is_race_other, data = dem_mod_dat)

  # Random effects: intercept + selected covariates
  Z <- model.matrix(~ is_female + is_married + comorbidity +
                      num_e4_1 + num_e4_2 + is_race_black + is_race_other,
                    data = dem_mod_dat)

  # Add random intercept as first column
  Z <- cbind(1, Z)

  # Map subjects to 0-based index for TMB
  id <- as.integer(as.factor(dem_mod_dat$adc)) - 1

  # Prepare data for TMB
  data <- list(
    y = dem_mod_dat$mod_y,
    X = X,
    Z = Z,
    id = id,
    L = dem_mod_dat$mod_l
  )

  # Starting values
  p <- ncol(X)
  q <- ncol(Z)
  n_subj <- length(unique(id))

  parameters <- list(
    beta = rep(0, p),
    log_sigma = 0,
    b_raw = rep(0, n_subj * q),
    log_sd_b = rep(0, q)
  )

  # Create objective function
  tic()
  obj <- MakeADFun(data, parameters, random = "b_raw", DLL = "truncated_lmm")

  # Optimize
  opt <- nlminb(obj$par, obj$fn, obj$gr)

  # Get standard errors
  rep <- sdreport(obj)
  toc()

  cat("\n\nPart 2 Random Effects Model Results:\n")
  print(summary(rep))

  # Store results
  part_2_re <- list(
    opt = opt,
    report = rep,
    obj = obj
  )

} else {
  cat("⚠ TMB model file not found at:", cpp_file, "\n")
  cat("  Skipping random effects model for Part 2\n")
  part_2_re <- NULL
}
```

### Table S.4 and S.5

Generate supplementary tables for ADRC analysis:

```{r adrc-tables}
# Table S.4: Part 1 Random Effects Summary
cat("\n\nTable S.4: Part 1 Model - Fixed Effects and Random Effects Variance\n")
cat("====================================================================\n\n")

# Fixed effects
fe_p1 <- fixef(part_1_re)
se_p1 <- sqrt(diag(vcov(part_1_re)))

table_s4 <- data.frame(
  Covariate = names(fe_p1),
  Estimate = round(fe_p1, 3),
  SE = round(se_p1, 3),
  Z = round(fe_p1 / se_p1, 3),
  P_value = format.pval(2 * pnorm(-abs(fe_p1 / se_p1)), digits = 3, eps = 0.001)
)

print(table_s4)

# Random effects variance
cat("\n\nRandom Effects Variance (ADRC) for Part 1:\n")
print(re_var_p1)

# Table S.5: Part 2 Random Effects Summary
cat("\n\nTable S.5: Part 2 Model - Fixed Effects and Random Effects Variance\n")
cat("====================================================================\n\n")

if (!is.null(part_2_re)) {
  # Extract results from TMB sdreport
  rep_summary <- summary(part_2_re$report, "fixed")

  # Beta coefficients (fixed effects)
  beta_idx <- grep("beta", rownames(rep_summary))
  beta_results <- rep_summary[beta_idx, ]

  # Create covariate names
  covariate_names <- c("(Intercept)", "t", "years_education", "is_female",
                      "is_married", "comorbidity", "num_e4_1", "num_e4_2",
                      "is_race_black", "is_race_other")

  table_s5 <- data.frame(
    Covariate = covariate_names,
    Estimate = round(beta_results[, "Estimate"], 3),
    SE = round(beta_results[, "Std. Error"], 3),
    Z = round(beta_results[, "Estimate"] / beta_results[, "Std. Error"], 3),
    P_value = format.pval(2 * pnorm(-abs(beta_results[, "Estimate"] /
                                         beta_results[, "Std. Error"])),
                         digits = 3, eps = 0.001)
  )

  print(table_s5)

  # Random effects standard deviations
  cat("\n\nRandom Effects Standard Deviations (ADRC) for Part 2:\n")
  sd_b_idx <- grep("log_sd_b", rownames(rep_summary))
  sd_b_results <- exp(rep_summary[sd_b_idx, "Estimate"])
  names(sd_b_results) <- c("Intercept", "is_female", "is_married", "comorbidity",
                          "num_e4_1", "num_e4_2", "is_race_black", "is_race_other")
  print(round(sd_b_results, 4))

  # Residual standard deviation
  cat("\n\nResidual Standard Deviation:\n")
  sigma_idx <- grep("log_sigma", rownames(rep_summary))
  sigma <- exp(rep_summary[sigma_idx, "Estimate"])
  cat("  σ =", round(sigma, 4), "\n")

} else {
  cat("⚠ Part 2 random effects model not available (TMB compilation failed)\n")
}

# Likelihood ratio test: Random effects vs. fixed effects
cat("\n\nLikelihood Ratio Test: Random vs. Fixed Effects\n")
cat("================================================\n\n")

cat("Part 1 (logistic regression):\n")
# Compare to fixed effects only model
lrt_p1 <- anova(part_1_mod, part_1_re)
print(lrt_p1)

if (!is.null(part_2_re)) {
  cat("\n\nPart 2 (truncated regression):\n")
  # Compare to basic truncated regression
  loglik_fixed <- logLik(fit_basic)
  loglik_random <- -part_2_re$opt$objective

  lr_stat <- 2 * (loglik_random - as.numeric(loglik_fixed))
  df_diff <- ncol(Z)  # Number of random effect standard deviations
  p_value <- pchisq(lr_stat, df = df_diff, lower.tail = FALSE)

  cat("Log-likelihood (fixed effects only):", round(as.numeric(loglik_fixed), 2), "\n")
  cat("Log-likelihood (random effects):", round(loglik_random, 2), "\n")
  cat("LR statistic:", round(lr_stat, 2), "\n")
  cat("df:", df_diff, "\n")
  cat("p-value:", format.pval(p_value, digits = 3, eps = 0.001), "\n")

  if (p_value < 0.05) {
    cat("\n✓ Random effects significantly improve model fit (p < 0.05)\n")
  } else {
    cat("\nℹ Random effects do not significantly improve fit (p ≥ 0.05)\n")
  }
}
```

## 6.2 Temporal Sampling Sub-Analysis

Use `death_year` to simulate historical analyses:

NOTE: This will take about 15 minutes to run

```{r temporal-analysis}
cat("Temporal Sampling Sensitivity Analysis\n")
cat("======================================\n\n")

if (!"death_year" %in% names(mod_dat)) {
  cat("⚠ death_year column not found in data\n")
  cat("  This analysis requires simulated death years\n")
} else {
  cat("Analyzing model stability across different observation periods...\n\n")

  # Define historical cutoff years
  # Include 2015 for Table S.7
  cutoff_years <- c(2010, 2015, 2020)

  temporal_results <- list()

  for (year in cutoff_years) {
    cat("Fitting models as if data collection stopped in", year, "...\n")

    # Filter data
    historical_data <- mod_dat %>% filter(death_year <= year)

    cat("  Sample size:", nrow(historical_data), "\n")

    # Fit Part 1
    part_1_historical <- glm(
      got_dementia ~ t + years_education + is_female + is_married + comorbidity +
        as.factor(num_e4) + is_race_black + is_race_other + l,
      data = historical_data,
      family = "binomial"
    )

    # Fit Part 2
    dem_historical <- historical_data %>%
      filter(got_dementia == 1) %>%
      mutate(num_e4_1 = ifelse(num_e4 == 1, 1, 0),
             num_e4_2 = ifelse(num_e4 == 2, 1, 0))

    if (nrow(dem_historical) > 50) {  # Need sufficient sample
      part_2_historical <- ltrc(
        survival::Surv(mod_y, got_dementia) ~ t + years_education + is_female +
          is_married + comorbidity + num_e4_1 + num_e4_2 + is_race_black + is_race_other,
        data = dem_historical,
        trunc_time = dem_historical$mod_l,
        n_start = 10,
        int_knots = 1
      )

      part_2_historical_clean <- get_clean_model(part_2_historical)

      # Calculate standard errors and CIs
      part1_se <- summary(part_1_historical)$coef[, 2]
      part1_lower <- coef(part_1_historical) - 1.96 * part1_se
      part1_upper <- coef(part_1_historical) + 1.96 * part1_se

      part2_se <- sqrt(diag(vcov(part_2_historical_clean)))
      part2_lower <- part_2_historical_clean$parameters$beta - 1.96 * part2_se
      part2_upper <- part_2_historical_clean$parameters$beta + 1.96 * part2_se

      # Store coefficients and CIs
      temporal_results[[as.character(year)]] <- list(
        year = year,
        n = nrow(historical_data),
        n_dementia = nrow(dem_historical),
        part1_coef = coef(part_1_historical),
        part2_coef = part_2_historical_clean$parameters$beta,
        part1_lower = part1_lower,
        part1_upper = part1_upper,
        part2_lower = part2_lower,
        part2_upper = part2_upper
      )
    }

    cat("\n")
  }

  # Compare coefficients over time
  cat("\nCoefficient Stability Analysis:\n")
  cat("================================\n\n")

  # Part 1 comparison
  cat("Part 1 Coefficients Across Years:\n")
  part1_comparison <- sapply(temporal_results, function(x) x$part1_coef[-1])  # Exclude intercept
  colnames(part1_comparison) <- sapply(temporal_results, function(x) x$year)
  print(round(part1_comparison, 3))

  # Part 2 comparison
  cat("\n\nPart 2 Coefficients Across Years:\n")
  part2_comparison <- sapply(temporal_results, function(x) x$part2_coef)
  colnames(part2_comparison) <- sapply(temporal_results, function(x) x$year)
  print(round(part2_comparison, 3))
}
```

### Figure S.6: Temporal Stability

```{r figure-s6, fig.width=10, fig.height=6}
if (exists("temporal_results") && length(temporal_results) > 0) {
  # Extract ε4 covariate results (one and two alleles)
  temporal_df <- data.frame()

  for (i in seq_along(temporal_results)) {
    res <- temporal_results[[i]]

    # Part 1: as.factor(num_e4)1 and as.factor(num_e4)2 (positions 7 and 8)
    # Part 2: num_e4_1 and num_e4_2 (positions 6 and 7)

    year_data <- data.frame(
      cutoff_year = as.Date(paste0(res$year, "-01-01")),
      model = rep(c("Part 1", "Part 2"), each = 2),
      label = rep(c("One ε4", "Two ε4"), 2),
      estimate = c(res$part1_coef[7:8], res$part2_coef[6:7]),
      ci_lower = c(res$part1_lower[7:8], res$part2_lower[6:7]),
      ci_upper = c(res$part1_upper[7:8], res$part2_upper[6:7])
    )

    temporal_df <- bind_rows(temporal_df, year_data)
  }

  # Create plot matching paper format
  fig_s6 <- temporal_df %>%
    ggplot() +
    geom_ribbon(
      aes(x = cutoff_year, ymin = ci_lower, ymax = ci_upper),
      fill = "gray", alpha = 0.5
    ) +
    geom_line(
      aes(x = cutoff_year, y = estimate),
      linewidth = 0.8
    ) +
    facet_grid(label ~ model, scales = "free_y") +
    labs(
      x = "Year Analysis was Run",
      y = "Estimated Effect"
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      text = element_text(family = "Times"),
      strip.background = element_blank(),
      strip.text = element_text(size = 12),
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12),
      panel.grid.minor = element_blank()
    )

  print(fig_s6)

  # Save
  ggsave(fs::path(RESULTS_DIR, "figure_s6_temporal_stability.png"),
         fig_s6, width = 10, height = 6, dpi = 300)
}
```

### Table S.7: Temporal Analysis (2015 Model Results)

```{r table-s7}
if (exists("temporal_results") && length(temporal_results) > 0) {
  # Extract 2015 results
  res_2015 <- temporal_results[["2015"]]

  if (!is.null(res_2015)) {
    cat("\n2015 Temporal Model Summary:\n")
    cat("  Sample Size:", res_2015$n, "\n")
    cat("  Dementia Cases:", res_2015$n_dementia, "\n")
    cat("  % Dementia:", round(res_2015$n_dementia / res_2015$n * 100, 1), "%\n\n")

    # Part 1: Extract coefficients (excluding intercept)
    part1_coef <- res_2015$part1_coef[-1]
    part1_lower <- res_2015$part1_lower[-1]
    part1_upper <- res_2015$part1_upper[-1]
    part1_se <- (part1_upper - part1_lower) / (2 * 1.96)
    part1_z <- part1_coef / part1_se

    # Part 2: Extract coefficients
    part2_coef <- res_2015$part2_coef
    part2_lower <- res_2015$part2_lower
    part2_upper <- res_2015$part2_upper
    part2_se <- (part2_upper - part2_lower) / (2 * 1.96)
    part2_z <- part2_coef / part2_se

    # Create table matching Table 2 format
    table_s7 <- data.frame(
      Covariate = c("t (time since age 65)", "Years of education", "Female",
                    "Married", "≥1 Comorbidity", "1 ε4 allele", "2 ε4 alleles",
                    "Black race", "Other race", "l (left truncation)"),
      Part1_Est = round(part1_coef, 3),
      Part1_CI = paste0("(", round(part1_lower, 3), ", ", round(part1_upper, 3), ")"),
      Part1_P = format.pval(2 * pnorm(-abs(part1_z)), digits = 3, eps = 0.001),
      Part2_Est = c(round(part2_coef, 3), "—"),
      Part2_CI = c(paste0("(", round(part2_lower, 3), ", ", round(part2_upper, 3), ")"), "—"),
      Part2_P = c(format.pval(2 * pnorm(-abs(part2_z)), digits = 3, eps = 0.001), "—")
    )

    knitr::kable(
      table_s7,
      caption = "Table S.7: Two-Part Model Coefficients (2015 Temporal Cutoff)",
      col.names = c("Covariate", "Estimate", "95% CI", "P-value",
                    "Estimate", "95% CI", "P-value"),
      align = c("l", "r", "l", "r", "r", "l", "r")
    )
  } else {
    cat("⚠ 2015 results not found in temporal_results.\n")
    cat("  Make sure cutoff_years includes 2015.\n")
  }
} else {
  cat("⚠ Temporal analysis results not available.\n")
}
```

## 6.3 Weighted Analysis

Fit models with inverse probability weighting:

```{r weighted-analysis}
weights_3 <- list(
  num_e4 = 0:2,
  is_race_black = 0:1,
  is_race_other = 0:1,
  is_female = 0:1
) %>%
  expand.grid() %>%
  filter(!(is_race_black == 1 & is_race_other == 1)) %>%
  arrange(num_e4, is_race_black, is_race_other, is_female) %>%
  mutate(
    obs_props = mod_dat %>% group_by(num_e4, is_race_black, is_race_other, is_female) %>% summarize(n = n(), pct = n() / nrow(.)) %>% pull(pct),
    pop_props = c(
      0.267784, # num_e4: 0 | race: white | sex: male
      0.267784, # num_e4: 0 | race: white | sex: female
      0.04117, # num_e4: 0 | race: other | sex: male
      0.04117, # num_e4: 0 | race: other | sex: female
      0.049046, # num_e4: 0 | race: black | sex: male
      0.049046, # num_e4: 0 | race: black | sex: female
      0.097614, # num_e4: 1 | race: white | sex: male
      0.097614, # num_e4: 1 | race: white | sex: female
      0.0150075, # num_e4: 1 | race: other | sex: male
      0.0150075, # num_e4: 1 | race: other | sex: female
      0.0178785, # num_e4: 1 | race: black | sex: male
      0.0178785, # num_e4: 1 | race: black | sex: female
      0.008602, # num_e4: 2 | race: white | sex: male
      0.008602, # num_e4: 2 | race: white | sex: female
      0.0013225, # num_e4: 2 | race: other | sex: male
      0.0013225, # num_e4: 2 | race: other | sex: female
      0.0015755, # num_e4: 2 | race: black | sex: male
      0.0015755  # num_e4: 2 | race: black | sex: female
    ),
    weight = pop_props / obs_props
  ) %>%
  select(-obs_props, -pop_props)


mod_dat <- mod_dat %>%
  left_join(weights_3, by = c("num_e4", "is_race_black", "is_race_other", "is_female"))

part_1_mod <- glm(
  got_dementia ~ t + years_education + is_female + is_married + comorbidity + is_race_black + is_race_other + as.factor(num_e4) + l,
  data = mod_dat,
  family = "binomial",
  weights = mod_dat$weight
)
summary(part_1_mod)
# note that standard errors are not correct here, need to use sandwich estimator if you want to do inference
vcov_robust <- vcovHC(part_1_mod, type = "HC0")
coeftest(part_1_mod, vcov = vcov_robust)

dem_mod_dat <- mod_dat %>%
  filter(got_dementia == 1) %>%
  mutate(
    num_e4_1 = ifelse(num_e4 == 1, 1, 0),
    num_e4_2 = ifelse(num_e4 == 2, 1, 0)
  )

tictoc::tic()
part_2_mod <- ltrc(
  survival::Surv(mod_y, got_dementia) ~ t + years_education + is_female + is_married + comorbidity + is_race_black + is_race_other + num_e4_1 + num_e4_2,
  data = dem_mod_dat, trunc_time = dem_mod_dat$mod_l, n_start = 10, int_knots = 1, weights = dem_mod_dat$weight
)
tictoc::toc()

part_2_mod_clean <- get_clean_model(part_2_mod)
part_2_mod_clean$parameters$beta
vcov(part_2_mod_clean) %>% diag() %>% sqrt()

readr::write_rds(part_2_mod_clean, fs::path(DATA_DIR, "weighted_pt_2_mod.rds"))

```

### Table S.8: Weighted Analysis Results

```{r}
cat("Part 1 Model - Weighted Analysis Results\n")
# CI
beta <- coef(part_1_mod)
se <- sqrt(diag(vcov_robust))
cbind(
  est = beta,
  lower = beta - 1.96 * se,
  upper = beta + 1.96 * se
)

cat("\n\nPart 2 Model - Weighted Analysis Results\n")
test_mod <- readr::read_rds(fs::path(DATA_DIR, "weighted_pt_2_mod.rds"))
beta <- test_mod$parameters$beta
se <- sqrt(diag(vcov(test_mod)))
cbind(
  est = beta,
  lower = beta - 1.96 * se,
  upper = beta + 1.96 * se
)
```

---

# 7. Simulation Study

This section demonstrates the simulation methodology used in the paper to validate the GHI estimation procedure. The simulation follows the exact code from `01_code_for_simulation_on_server.R`, omitting the bootstrap variance calculation since it is too time intensive. That code can be seen in the original file, and run following the instructions in the section on running the simulation on a server.

## Simulation Setup

```{r simulation-setup}
cat("GHI Model Validation via Simulation\n")
cat("====================================\n\n")

# Set true parameter values (used in paper)
alpha <- c(1.7, 0.05, 0.05)    # For generating t: log(t) ~ alpha[1] + alpha[2]*x_1 + alpha[3]*x_2
beta <- c(-0.1, -0.5, 2, 0.2)  # Part 1 coefficients: t, x_1, x_2, l
omega <- c(0, -0.1, 1, -1)     # Part 2 coefficients: intercept, t, x_1, x_2

cat("True parameters:\n")
cat("  α (t generation):", alpha, "\n")
cat("  β (Part 1 model):", beta, "\n")
cat("  ω (Part 2 model):", omega, "\n\n")

# Define prediction grid for GHI estimates
covariate_grid <- list(
  t = 7:12,                          # Age at death - 65
  x_1 = seq(-1.5, 1.5, by = 0.25),  # Continuous covariate
  x_2 = c(0, 1),                     # Binary covariate
  l = 1                              # Left truncation (age at entry - 65)
) %>%
  expand.grid() %>%
  mutate(
    l_over_t = l / t,
    eps_lower = log(l_over_t / (1 - l_over_t)) - omega[2] * t - omega[3] * x_1 - omega[4] * x_2
  )

cat("Prediction grid:\n")
cat("  t values:", range(covariate_grid$t), "\n")
cat("  x_1 values:", range(covariate_grid$x_1), "\n")
cat("  x_2 values:", unique(covariate_grid$x_2), "\n")
cat("  Total grid points:", nrow(covariate_grid), "\n\n")
```

## Simulation Function

```{r simulation-function}
# Main simulation function (matches server code exactly)
ghi_sim_run <- function(i, sample_size, error_fn = rnorm, error_name = "Standard Normal",
                        alpha = c(1.7, 0.05, 0.05),
                        beta = c(-0.1, -0.5, 2, 0.2),
                        omega = c(0, -0.1, 1, -1),
                        covariate_grid) {

  cat("Iteration", i, "\n")

  # Step 1: Generate true data (before truncation/selection)
  true_dat <- data.frame(
    l = runif(4 * sample_size, 0, 5),      # Entry age - 65
    x_1 = runif(4 * sample_size, -3, 3),   # Continuous covariate
    x_2 = rbinom(4 * sample_size, 1, 0.4)  # Binary covariate
  ) %>%
    mutate(
      # Generate t (age at death - 65)
      t = exp(alpha[1] + alpha[2] * x_1 + alpha[3] * x_2 + rnorm(4 * sample_size, 0, 0.3)),
      # Generate error term
      eps = error_fn(4 * sample_size),
      # Generate y (proportional dementia timing)
      y = exp(omega[1] + omega[2] * t + omega[3] * x_1 + omega[4] * x_2 + eps) /
          (1 + exp(omega[1] + omega[2] * t + omega[3] * x_1 + omega[4] * x_2 + eps)),
      # Dementia age (s = y*t)
      s = y * t
    )

  # Step 2: Apply left-truncation (keep only s > l and t > l)
  obs_dat <- true_dat %>%
    filter(s > l, t > l) %>%
    mutate(
      # Generate dementia indicator from Part 1 model
      prob_disease = exp(beta[1] * t + beta[2] * x_1 + beta[3] * x_2 + beta[4] * l) /
                     (1 + exp(beta[1] * t + beta[2] * x_1 + beta[3] * x_2 + beta[4] * l)),
      delta_s = rbinom(nrow(.), 1, prob_disease),
      # Set y=1 for non-dementia cases
      y = ifelse(delta_s, y, 1)
    )

  # Step 3: Sample down to target size
  sim_dat <- obs_dat %>% sample_n(size = sample_size)

  # Step 4: Fit Part 1 model
  part_1_mod <- glm(delta_s ~ t + x_1 + x_2 + l, data = sim_dat, family = binomial)

  # Step 5: Fit Part 2 model (on dementia cases only)
  got_s <- sim_dat %>%
    filter(delta_s == 1) %>%
    mutate(
      mod_y = log(y / (1 - y)),
      scaled_l = l / t,
      mod_l = log(scaled_l / (1 - scaled_l))
    )

  part_2_mod <- tryCatch({
    ltrc(survival::Surv(mod_y, delta_s) ~ t + x_1 + x_2,
         data = got_s,
         trunc_time = got_s$mod_l,
         int_knots = 2)  # Using 2 interior knots
  }, error = function(e) {
    print(e)
    return(NULL)
  })

  if (is.null(part_2_mod)) return(NULL)

  part_2_mod_clean <- get_clean_model(part_2_mod)
  beta_hat <- part_1_mod$coefficients[-1]
  omega_hat <- part_2_mod_clean$parameters$beta
  part_2_residuals <- c(part_2_mod_clean$data$residuals)

  # Step 6: Generate predictions across covariate grid
  pred_df <- data.frame()
  lps <- covariate_grid$t * omega_hat[1] + covariate_grid$x_1 * omega_hat[2] +
         covariate_grid$x_2 * omega_hat[3]

  for (j in 1:nrow(covariate_grid)) {
    # Part 1: Probability of NOT getting dementia
    mu_0 <- 1 - predict(part_1_mod, newdata = covariate_grid[j, 1:4], type = "response")

    # Part 2: Expected Y given dementia via Kaplan-Meier
    sub_lp <- lps[j]
    cutoff <- covariate_grid[j, 6]
    tau <- got_s %>%
      mutate(
        l_resid_scale = mod_l - omega_hat[1] * t - omega_hat[2] * x_1 - omega_hat[3] * x_2
      ) %>%
      pull(l_resid_scale)

    km_df <- data.frame(epsilon = part_2_residuals, tau = tau, delta = 1) %>%
      mutate(
        g_of_e = exp(sub_lp + epsilon) / (1 + exp(sub_lp + epsilon)),
        g_of_t = exp(sub_lp + tau) / (1 + exp(sub_lp + tau))
      ) %>%
      filter(epsilon > cutoff)

    km_fit <- tryCatch({
      survfit(Surv(time = g_of_t, time2 = g_of_e, event = delta) ~ 1, data = km_df)
    }, error = function(e) NULL)

    if (is.null(km_fit)) next

    km_times <- c(0, km_fit$time)
    km_surv <- c(1, km_fit$surv)
    km_estimate <- sum(km_surv[-length(km_surv)] * diff(km_times))

    # GHI = (1-mu) + mu * E[Y | dementia]
    area_expectation <- mu_0 + (1 - mu_0) * km_estimate

    # Variance calculation
    g_n <- mu_0
    x_mat <- matrix(c(1, as.numeric(covariate_grid[j, 1:4])), ncol = 1)
    sigma_11 <- (g_n * (1 - g_n))^2 * t(x_mat) %*% vcov(part_1_mod) %*% x_mat

    tmp_preds <- data.frame(
      t = covariate_grid[j, 1],
      x_1 = covariate_grid[j, 2],
      x_2 = covariate_grid[j, 3],
      l = covariate_grid[j, 4],
      mu_0 = mu_0,
      expected_y = km_estimate,
      ghi = area_expectation,
      sigma_11 = sigma_11
    )

    pred_df <- rbind(pred_df, tmp_preds)
  }

  # Return results
  list(
    iter_num = i,
    n = sample_size,
    pct_truncation = 1 - nrow(obs_dat) / nrow(true_dat),
    error_name = error_name,
    beta = beta,
    omega = omega,
    beta_hat = beta_hat,
    omega_hat = omega_hat,
    data = sim_dat,
    part_1_model = part_1_mod,
    part_2_model = part_2_mod_clean,
    prediction_df = pred_df
  )
}

cat("✓ Simulation function defined\n")
```

## Local Simulation (Small Scale)

```{r simulation-local, eval=FALSE}
cat("\nRunning Small-Scale Demonstration\n")
cat("==================================\n\n")

SAMPLE_SIZE <- 400
N_SIM <- 5

cat("⚠ This is a minimal demonstration\n")
cat("  Sample size:", SAMPLE_SIZE, "\n")
cat("  Replications:", N_SIM, "\n")
cat("  For production results, use server scripts with n=400, R=300 bootstrap\n\n")

big_result <- list()

tictoc::tic()
for (m in 1:N_SIM) {
  result <- ghi_sim_run(
    i = m,
    sample_size = SAMPLE_SIZE,
    error_fn = rnorm,
    error_name = "Standard Normal",
    alpha = alpha,
    beta = beta,
    omega = omega,
    covariate_grid = covariate_grid
  )

  if (!is.null(result)) {
    big_result <- append(big_result, list(result))
  }
}
tictoc::toc()

cat("\n✓ Simulation complete\n")
cat("  Successful runs:", length(big_result), "/", N_SIM, "\n")

# Summary of results
if (length(big_result) > 0) {
  cat("\nParameter Recovery:\n")
  cat("===================\n\n")

  # Extract estimates
  beta_estimates <- t(sapply(big_result, function(x) x$beta_hat))
  omega_estimates <- t(sapply(big_result, function(x) x$omega_hat))

  cat("Part 1 (β) - True vs Estimated (mean):\n")
  comparison_beta <- data.frame(
    Parameter = c("t", "x_1", "x_2", "l"),
    True = beta,
    Mean_Est = colMeans(beta_estimates),
    Bias = colMeans(beta_estimates) - beta,
    SD = apply(beta_estimates, 2, sd)
  )
  print(comparison_beta)

  cat("\nPart 2 (ω) - True vs Estimated (mean):\n")
  cat("  Note: ltrc model does not estimate intercept (fixed at 0)\n\n")
  comparison_omega <- data.frame(
    Parameter = c("t", "x_1", "x_2"),
    True = omega[2:4],  # Skip intercept (omega[1] = 0, not estimated)
    Mean_Est = colMeans(omega_estimates),
    Bias = colMeans(omega_estimates) - omega[2:4],
    SD = apply(omega_estimates, 2, sd)
  )
  print(comparison_omega)

  cat("\nTruncation rates:\n")
  trunc_rates <- sapply(big_result, function(x) x$pct_truncation)
  cat("  Mean:", round(mean(trunc_rates), 3), "\n")
  cat("  Range:", round(range(trunc_rates), 3), "\n")

  # Aggregate prediction results across covariate grid
  cat("\n\nPrediction Grid Results:\n")
  cat("========================\n\n")

  # Combine all prediction dataframes
  all_preds <- bind_rows(lapply(1:length(big_result), function(i) {
    big_result[[i]]$prediction_df %>%
      mutate(iter = i)
  }))

  # Aggregate by covariate values
  pred_summary <- all_preds %>%
    group_by(t, x_1, x_2, l) %>%
    summarize(
      n_reps = n(),
      mean_mu_0 = mean(mu_0, na.rm = TRUE),
      mean_expected_y = mean(expected_y, na.rm = TRUE),
      mean_ghi = mean(ghi, na.rm = TRUE),
      sd_ghi = sd(ghi, na.rm = TRUE),
      se_ghi = sd(ghi, na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    ) %>%
    arrange(t, x_1, x_2)

  cat("Total grid points:", nrow(pred_summary), "\n")
  cat("Replications per point:", unique(pred_summary$n_reps), "\n\n")

  # Display sample of results (first 20 rows)
  cat("Sample of prediction grid (first 20 rows):\n\n")
  pred_display <- pred_summary %>%
    select(t, x_1, x_2, mean_mu_0, mean_expected_y, mean_ghi, se_ghi) %>%
    head(20)

  print(knitr::kable(
    pred_display,
    digits = 4,
    col.names = c("t", "x_1", "x_2", "P(No Dem)", "E[Y|Dem]", "GHI", "SE(GHI)"),
    caption = "Prediction Grid Results Summary (Sample)"
  ))

  # Summary statistics across grid
  cat("\n\nSummary across entire grid:\n")
  cat("  GHI range:", round(range(pred_summary$mean_ghi), 4), "\n")
  cat("  Mean GHI:", round(mean(pred_summary$mean_ghi), 4), "\n")
  cat("  Mean SE(GHI):", round(mean(pred_summary$se_ghi), 4), "\n")
}
```

## Large-Scale Simulation (Server Instructions)

The production simulation used the following specification:
- **Sample size**: n = 400, 800, or 1600 per replication
- **Replications**: N_SIM = 5 per file
- **Bootstrap samples**: 300 per replication (for variance estimation)
- **Total files**: 50 (yielding 250 total replications)

```markdown
# Large-Scale Simulation on Server

## Setup

1. Upload scripts to server:
   - `01_code_for_simulation_on_server.R` (main simulation code)
   - `02_run_simulation_parallel.sh` (parallel execution script)

2. Create output directory:
   ```bash
   mkdir -p ghi_sim_res
   ```

3. Review parameters in `01_code_for_simulation_on_server.R`:
   - `N_SIM = 5`: Number of simulation runs per file
   - `SAMPLE_SIZE = 400`: Sample size per replication (400, 800, or 1600)
   - `n_bootstrap = 300`: Bootstrap samples for variance (optional, set to 0 to skip)
   - True parameters: α, β, ω as defined in simulation setup

## Running

### Option A: Bash Script (Background Jobs)

```bash
cd inst/paper_code
bash 02_run_simulation_parallel.sh
wait  # Wait for all jobs to complete
```

### Option B: Manual Parallel Execution

```bash
# Run 50 files in parallel (FILE_NUM = 1 to 50)
for i in {1..50}; do
  Rscript 01_code_for_simulation_on_server.R $i > out_$i.txt 2>&1 &
done
wait
```

## Expected Runtime

- **Per file** (5 replications + 300 bootstrap each): 2-4 hours
- **Total (50 parallel jobs)**: 2-4 hours wall time
- **Total (sequential)**: 100-200 hours

## Output

- 50 `.rds` files in `ghi_sim_res/`
- Each file: `error_a_400_{FILE_NUM}.rds`
- Each contains 5 simulation results with:
  - Parameter estimates (β, ω)
  - Prediction dataframes across covariate grid
  - Bootstrap variance estimates (if n_bootstrap > 0)
  - Truncation rates

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# Summary and Next Steps

## What We Did

- *Setup**: Configured environment and loaded simulated NACC data
- *Data Exploration**: Created Table 1 with descriptive statistics
- *Model Fitting**: Fit two-part model (GLM + LTRC)
- *Model Analysis**: Generated Table 2 and Figures 1-2
- *Model Checking**: Comprehensive diagnostics for both model parts
- *Sensitivity Analyses**: ADRC, temporal, and weighted analyses
- *Simulation Study**: Demonstrated methodology (small-scale)

## For Reviewers

This notebook provides a complete walkthrough of our methodology using simulated data. Key points:

1. **All code is transparent** - every analysis can be inspected and modified
2. **Simulated data mimics real NACC structure** - methods are identical
3. **Results differ from paper** - expected due to simulated vs. real data
4. **Computationally intensive parts** - marked with warnings, can be run on server

## Reproducing Paper Results

To reproduce **exact** results from the paper:

1. Obtain real NACC data (requires data use agreement)
2. Run `03_create_NACC_model_data.R` to prepare data
3. Execute remaining scripts in order (see Paper_Reproduction_NACC.md.md)
4. For bootstrap variance: use server scripts (06, 07)
5. For simulation study: use server scripts (01, 02)

## Questions?

For questions about this analysis or the methods, please contact the authors at information provided in the paper.

---

**Document generated:** `r Sys.time()`
